A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-09 doc/popl2013/intro.tex
intro

In Legion data is organized in a hierarchy of {\em regions}
and subregions while computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region permissions} specifying 
operations a task may perform on a region argument (read,
write, or reduce) and novel {\em region coherence} annotations that
express what other tasks may do concurrently with
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the permissions and coherence system and use these two
results to prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em independent} (have no ordering requirements for correctness), 
then the unique ancestors of $t_1$ and $t_2$ in the task hierarchy 
are also independent.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in communicating in the target
class of machines.

% The linked list example
% Add float={t} to the list of options if you want the listing at the top
\begin{lstlisting}[float={t},label={lst:linkedlist},caption={Linked List Example}]
type list<r> = < int, list<r>@r >

task sum_list[r] ( l:list<r>@r ), reads(r) : int =
    if isnull(l) then 0 else
    let v : list<r> = read(l) in
        v.1 + sum_list[r](v.2)
\end{lstlisting}

Legion programs organize data into {\em logical regions}.  Logical
regions (which we usually refer to simply as {\em regions}) express
{\em locality}, which is simply data that will be used together.
Listing~\ref{lst:linkedlist} gives an example that defines a region
containing a linked list of integers and a task that sums the elements
of the list.  As is standard in region systems, we use $T @ r$ for the
type of a pointer to a value of type $T$ in region $r$.  The task also
illustrates a simple use of the permissions system:
task {\tt sum\_list} must have permission {\tt reads(r)}, the ability
to read the values in region {\tt r}.

>>>>

In Legion data is organized in a hierarchy of {\em regions}
and subregions while computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region permissions} specifying 
operations a task may perform on a region argument (read,
write, or reduce) and  {\em region coherence} annotations that
express what other tasks may do concurrently with
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the permissions and coherence system and use these two
results to prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em independent} (have no ordering requirements for correctness), 
then the unique ancestors of $t_1$ and $t_2$ in the task hierarchy 
are also independent.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in communicating in the target
class of machines.

\lstset{
  captionpos=b,
  language=C++,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={region,coloring,partition,spawn,disjoint,aliased},
  deletekeywords=float,
}
\begin{lstlisting}[float={t},label={lst:code_ex},caption={Circuit simulation.}]
struct Node { float voltage, new_charge, capacitance; };
struct Wire<rn> { Node@rn in_node, out_node; float current, ... ; };
struct Circuit { region   r_all_nodes; /* contains all nodes for the circuit */
                 region   r_all_wires; /* contains all circuit wires */ };
struct CircuitPiece {
  region  rn_pvt, rn_shr, rn_ghost; /* private, shared, ghost node regions */
  region  rw_pvt;                   /* private wires region */ };

void simulate_circuit(Circuit c, float dt) : RWE(c.r_all_nodes, c.r_all_wires)
{
  // The construction of the colorings is not shown.  The colorings wire_owner_map,
  // node_owner_map, and node_neighbor_map have MAX_PIECES colors 
  // 0..MAX_PIECES #$-$# 1. The coloring node_sharing map has two colors 0 and 1.
  //
  // Partition of wires into MAX_PIECES pieces
  partition<disjoint> p_wires = c.r_all_wires.partition(wire_owner_map); 
  // Partition nodes into two parts for all-private vs. all-shared
  partition<disjoint> p_nodes_pvs = c.r_all_nodes.partition(node_sharing map);

  // Partition all-private into MAX_PIECES disjoint circuit pieces
  partition<disjoint> p_pvt_nodes = p_nodes_pvs[0].partition(node_owner_map);
  // Partition all-shared into MAX_PIECES disjoint circuit pieces
  partition<disjoint> p_shr_nodes = p_nodes_pvs[1].partition(node_owner_map);
  // Partition all-shared into MAX_PIECES ghost regions, which may be aliased
  partition<aliased> p_ghost_nodes = p_nodes_pvs[1].partition(node_neighbor_map);

  CircuitPiece pieces[MAX_PIECES];
  for(i = 0; i #$<$# MAX_PIECES; i++) 
    pieces[i] = { rn_pvt: p_pvt_nodes[i], rn_shr: p_shr_nodes[i],
                  rn_ghost: p_ghost_nodes[i], rw_pvt: p_wires[i] };
  for (t = 0; t #$<$# TIME_STEPS; t++) {
    for (i = 0; i #$<$# MAX_PIECES; i++) calc_new_currents(pieces[i]);
    for (i = 0; i #$<$# MAX_PIECES; i++) distribute_charge(pieces[i], dt);
    for (i = 0; i #$<$# MAX_PIECES; i++) update_voltages(pieces[i]);
  }
}

void calc_new_currents(CircuitPiece piece):
        RWE(piece.rw_pvt), ROE(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
  foreach(w : piece.rw_pvt)
    w#$\rightarrow$#current = (w#$\rightarrow$#in_node#$\rightarrow$#voltage - w#$\rightarrow$#out_node#$\rightarrow$#voltage) / w#$\rightarrow$#resistance;
}

void distribute_charge(CircuitPiece piece, float dt):
        ROE(piece.rw_pvt), RdA(piece.rn_pvt, piece.rn_shr, piece.rn_ghost) {
  foreach(w : piece.rw_pvt) {
    w#$\rightarrow$#in_node#$\rightarrow$#new_charge += -dt * w#$\rightarrow$#current;
    w#$\rightarrow$#out_node#$\rightarrow$#new_charge +=  dt * w#$\rightarrow$#current;
  }
}

void update_voltages(CircuitPiece piece): RWE(piece.rn_pvt, piece.rn_shr) {
  foreach(n : piece.rn_pvt, piece.rn_shr) {
    n#$\rightarrow$#voltage += n#$\rightarrow$#new_charge / n#$\rightarrow$#capacitance;
    n#$\rightarrow$#new_charge = 0;
  }
}
\end{lstlisting}

Legion programs organize data into a forest of {\em logical regions}.  Logical
regions express {\em locality} (data that will be used together, and therefore may be colocated in the same
physical memory) and {\em independence} (disjoint data that may be operated on in parallel).
Beginning with Sequoia \cite{Fatahalion06} and Deterministic Parallel Java \cite{Bocchino09}, researchers
have explored hierarchical decomposition of data to express locality and independence.  Two

containing a linked list of integers and a task that sums the elements
of the list.  As is standard in region systems, we use $T @ r$ for the
type of a pointer to a value of type $T$ in region $r$.  The task also
illustrates a simple use of the permissions system:
task {\tt sum\_list} must have permission {\tt reads(r)}, the ability
to read the values in region {\tt r}.
